Limitations on the Two-Infall Scenario in the Era of Large Stellar Age Catalogs

### 2025-05-12 src/tex/ms.tex
Updates from Overleaf

Figure \ref{fig:ofe-feh-best} presents the stellar [O/Fe]--[Fe/H]--age distributions as a function of $R_{\rm gal}$ and $|z|$ predicted by the best multi-zone model.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/lmr_ages.pdf}
    \caption{{\it Top:} The median stellar age as a function of [O/Fe] and [Fe/H] in the Solar annulus ($7\leq R_{\rm gal}<9\kpc$, $0\leq|z|<2\kpc$). The left and center panels plot the output of our best two-infall models, with \yZ{2}, ${\rm [X/H]}_{\rm CGM}=-0.7$, $f_\Sigma(R_\odot)=0.25$, and $\eta_\odot=1.8$. The model output has been re-sampled to match the APOGEE stellar $|z|$ distribution, and a Gaussian scatter has been applied to the abundances and ages according to Table \ref{tab:uncertainties}. The right panel plots the results from APOGEE using the \citet{leung_variational_2023} age catalog. The contours indicate the density of stars in the [Fe/H]--[O/Fe] plane, and the vertical dashed line denotes the boundary for locally metal-rich (LMR) stars.
    {\it Bottom}: Stellar age distributions in the Solar annulus for all stars (black) and LMR stars (gray). The left panel plots the results from the two-infall model after forward-modeling age uncertainties, and the right panel plots the \citet{leung_variational_2023} ages for APOGEE stars.
    {\bf Key takeaway:} The two-infall model predicts a fundamentally different age pattern than what is observed, especially for LMR stars.}
    \label{fig:lmr-ages}
\end{figure*}

The two-infall model makes a prediction about the local stellar age distribution that is fundamental to its construction: that the most metal-rich stars in any region of the Galaxy come from the metal-rich tail of the first infall sequence, and are therefore older than all of the thin disk stars. This prediction can be seen in any of the panels in Figure \ref{fig:ofe-feh-best}. In Figure \ref{fig:lmr-ages}, there is a sharp divide in the median age of local metal-rich (LMR; $\mathFeH\gtrsim+0.1$) stars when moving from the thick disk ($\tau\ge9\Gyr$) to the thin disk ($\tau\lesssim5\Gyr$).

>>>>

Figure \ref{fig:ofe-feh-best} presents the stellar [O/Fe]--[Fe/H]--age distributions as a function of $R_{\rm gal}$ and $|z|$ predicted by the best multi-zone model. The model is generally successful at reproducing the observed distribution of stars in the [O/Fe]--[Fe/H] plane, especially in the inner Galaxy and close to the midplane (panels along the left and bottom sides of the figure, respectively). However, the predicted high-$\alpha$ sequence is less concentrated than in the data, and its presence is still significant even in the outer Galaxy---likely a consequence of the stronger migration prescription and higher thick-to-thin disk ratio. In general, the predicted distributions do not align with the data quite as well at large midplane distances ($1\leq|z|<2\kpc$), but this may partly be due to our prescription for vertical heating (see Section \ref{sec:migration}).

The model makes two notable predictions about the age--abundance distributions. First, there is a population of $\sim8-9\Gyr$ old stars at sub-Solar [O/Fe], especially at $|z|\ge0.5\kpc$, formed immediately after the second infall during a period of rapid chemical evolution. These stars form a small percentage of the overall distribution \citep[see also Figure 11 from][]{spitoni_remind_2024} but in this case they occupy a unique portion of the abundance space. A longer $\tau_1$ could shift this population to higher [O/Fe] where it would be obscured by the rest of the low-$\alpha$ sequence (see Figure \ref{fig:twoinfall-parameters}). Second, the stars born at the tail end of the thick and thin disk epochs are adjacent to each other in abundance space, meaning the two-infall model predicts a steep age gradient for the most metal-rich stars in a given region.

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/lmr_ages.pdf}
    \caption{{\it Top:} The median stellar age as a function of [O/Fe] and [Fe/H] in the Solar annulus ($7\leq R_{\rm gal}<9\kpc$, $0\leq|z|<2\kpc$). The left and center panels plot the output of our best two-infall models, with \yZ{2}, ${\rm [X/H]}_{\rm CGM}=-0.7$, $f_\Sigma(R_\odot)=0.25$, and $\sigma_{\rm RM8}=3.6\kpc$. The model output has been re-sampled to match the APOGEE stellar $|z|$ distribution, and a Gaussian scatter has been applied to the abundances and ages according to Table \ref{tab:uncertainties}. The right panel plots the results from APOGEE using the \citet{leung_variational_2023} age catalog. The contours indicate the density of stars in the [Fe/H]--[O/Fe] plane, and the vertical dashed line denotes the boundary for locally metal-rich (LMR) stars.
    {\it Bottom}: Stellar age distributions in the Solar annulus for all stars (black) and LMR stars (gray). The left and center panels plot the mass-weighted age distributions predicted by the models after forward-modeling age uncertainties, and the right panel plots the \citet{leung_variational_2023} ages for APOGEE stars.
    {\bf Key takeaway:} The two-infall model predicts a fundamentally different age pattern than what is observed, especially for LMR stars.}
    \label{fig:lmr-ages}
\end{figure*}

The two-infall model makes a prediction about the local stellar age distribution that is fundamental to its construction: that the most metal-rich stars in any region of the Galaxy come from the metal-rich tail of the first infall sequence, and are therefore older than all of the thin disk stars. As noted in the previous Section, this prediction is apparent in any of the panels in Figure \ref{fig:ofe-feh-best}, especially where $|z|<0.5\kpc$. We investigate this prediction further here.

The top row of Figure \ref{fig:lmr-ages} presents the median stellar age as a function of [O/Fe] and [Fe/H] for two multi-zone models and APOGEE. While the models predict a fairly accurate distribution of stars in abundance space, especially for the low-$\alpha$ population, the stellar age patterns are obviously quite different. In both the \yZ{1} and \yZ{2} models, there is a sharp divide in the median stellar age when moving from the thick disk ($\tau\ge9\Gyr$) to the thin disk ($\tau\lesssim5\Gyr$). The \yZ{2} model also predicts that the stars with the lowest [O/Fe] should be $\sim8-9\Gyr$ old, while these are some of the youngest stars in APOGEE. The latter issue can be mitigated by adjusting the parameters of the first infall, as discussed in the previous Section, but the former is not solved so easily.

We further highlight the discrepant age patterns in the bottom panels of Figure \ref{fig:lmr-ages}, which compare the overall stellar age distribution against that of the locally metal rich (LMR) stars, defined here as $\mathFeH\ge+0.1$.\footnote{The precise location of the cut matters little, as we observe the same behavior for cuts ranging from $+0.05$ to $+0.2$ dex.} For APOGEE, the distributions are similar, both peaking near $\sim5\Gyr$, although very few of the LMR stars have ages $\gtrsim10\Gyr$. Our two-infall models produce an age distribution for the overall sample that is similar to the data, but for LMR stars, both models predict a distinctly bimodal age distribution. There is a large contribution from the young, metal-rich end of the second infall, and a contribution from the old, metal-rich end of the first, but there are few stars in between. The trough between the modes lies at $\sim5\Gyr$ for both models, right where the APOGEE distribution peaks.

\todo{Mention some ways to try to resolve this (e.g., stronger radial migration) and that they don't work.}
\todo{Note that our projected log age error of 0.1 dex is accurate for stars $<8\Gyr$ old, but too large for older stars, so the scatter in the high-$\alpha$ sequence is larger than the data.}
